Fixing Vulnerabilities Potentially Hinders Maintainability

### 2020-05-28 paper/icsme20.tex
new version

\begin{abstract}
Security is a requirement of utmost importance to produce software with 
high quality. However, there is still a considerable amount of 
vulnerabilities being discovered and fixed, almost weekly. We suspect 
that when developers address these vulnerabilities on their codebases, 
they actually affect the maintainability of software, potentially 
introducing other vulnerabilities. This paper evaluates the impact of 
patches to improve security on the maintainability of open-source software. 
Maintainability is measured based on the Better Code Hub’s model of 10 
guidelines using a dataset including 1335 security commits.
Results show that fixing software vulnerabilities may hinder the maintainability 
of open-source software and that different levels of attention should be given to 
different severity levels, programming languages and types of weaknesses. We 
conclude that changes to code bases to patch 
vulnerabilities need to be performed with extra care and that security-related 
code evolution should be subject in computer curricula.\Rui{Needs improvement}
\end{abstract}

\begin{IEEEkeywords}
Security, Software Maintenance, Open-Source Software
\end{IEEEkeywords}

@@

 \begin{figure}[htp]
 	\vspace{-0.9cm}

     \begin{adjustwidth}{-1cm}{-1cm}  
  	\centering
  	\includegraphics[width=0.51\textwidth]{figures/main_per_guideline.pdf}

    \end{adjustwidth}
  	\caption{Impact of the security patches per guideline. For each guideline, $3$ types of impact in the software maintainability are presented: negative, maintainability decreases (red); none, maintainability remains the same (yellow); and, positive, maintainability increases (green).}
 	\label{fig:guidelines}	
	\vspace{-0.5cm}
 \end{figure}

Regarding the impact of security patches for each guideline, we observe
that overall patches have more cases where the impact is positive. However, 
we also see that patching vulnerabilities also has a very
significant negative impact on all guidelines---between $10\%$ and $40\%$. 
\emph{Write Short Units of 
Code} ($38.3\%$), \emph{Write Simple Units of Code} ($37.9\%$),  and 
\emph{Separate Concerns in Modules} ($33.8\%$) seem to be the 
most negativelly affected guidelines. 
This may imply that developers when patching 
vulnerabilities fail to design/implement solutions that continue to respect 
the limit bounds of branch points and function/module 
sizes that are recommended by coding best practices. Still, on 
respecting bound limits, developers also seem to not consider the limit 
of $4$ parameters per function for the \emph{Keep Unit Interfaces Small} 
guideline required by BCH, in $24.8\%$ of the cases. This guideline is usually
violated when the patch requires to input new information to a function/class 
and developers fail to use the \emph{Introduce Parameter Object} refactoring 
pattern. Results do not provide statistical significance to the \emph{Separate 
Concerns in Modules} guideline, i.e., results should be read carefully.

Software architecture is also affected while patching vulnerabilities.
Both \emph{Couple Architecture Component Loosely} and \emph{Keep
Architecture Components Balanced} guidelines suffer a negative impact of 
$26.9\%$ and $11.0\%$, respectively. Component independence and balance
are important to make easier to find the source code that developers
want to patch/improve and to understand how the high-level components
interact with others. However, results may imply that developers
fail to use techniques such as encapsulation to hide implementation
details and make the system more modular.

Software systems typically have $9\%$-$17\%$ of cloned code~\cite{5773403}. 
The \emph{Write Code Once} guideline results show that duplicated code 
increased in $171$ ($17.7\%$) of the patches. Previous work showed a 
correlation between code smells and code duplication~\cite{7476787} 
which may also be reflected in the \emph{Write Clean Code} guideline results. 
BCH reported new code smells for $145$ ($15.0\%$) patches which according 
to previous work may harm software security since some of these new code
smells may be new vulnerabilities~\cite{8819456}.
Thus, while developers 
tend to ignore refactoring techinques such as the \emph{Replace Method 
with Method Object} technique, they may also be introducing new software 
vulnerabilities and, consequently, harm the market value and economy of companies
~\cite{4267025}.

\textbf{Overall Score ($M(v)$):} 
Although overall patching vulnerabilities has a less negative impact
on software maintainability guidelines, this is not reflected in the average 
impact of all guidelines ($M(v)$) as we can see in Figure~\ref{fig:guidelines}.
The $M(v)$ plots show that $406$ ($41.9\%$) cases have negative impact
on software maintainability. While $188$ ($19.4\%$) have no impact and $375$ ($38.7\%$) cases
have a negative impact. The larger number of negative cases may be 
explained by guidelines with higher concentration of negative cases with higher amplitudes, 
such as, \emph{Write Short Units of 
Code}, \emph{Write Simple Units of 
Code} and \emph{Separate Concerns in Modules}.

@@

\textbf{Severity:} Some of the vulnerabilities are identified with \emph{Common 
Vulnerabilities and Exposure} (CVE) entries. We leveraged the \emph{National Vulnerability 
Database} (NVD) website to collect their severity levels. In total, we 
retrieved severity scores for $536$ vulnerabilities: $112$ (\emph{HIGH}), 
$395$ (\emph{MEDIUM}) and $29$ (\emph{LOW}). Figure~\ref{fig:severity} 
presents the impact of security patches per severity level on the 
maintainability of open-source software. We observe that patches for 
\emph{HIGH} ($50.0\%$) and \emph{MEDIUM} ($43.8\%$) severity vulnerabilities 
hinder more the maintainability of software than \emph{LOW} ($27.6\%$) 
severity vulnerabilities. Again, patches have a considerable 
negative impact on software maintainability---between $20\%$ 
and $50\%$. Statistical significance was retrieved only for \emph{LOW}
severity vulnerabilitie, i.e., \emph{LOW} severity vulnerabilities
may have less negative impact on software maintainability thant the 
other severity levels. However, 
results should not be disregarded because they seem to confirm the 
assumption that higher severity vulnerabilities patches may have a 
more negative impact on maintainability and, consequently, that high/medium 
severity vulnerabilities may need more attention than low severity while 
patching vulnerabilities.

@@

\textbf{Programming Language:} The impact on software maintainability per programming 
language was also analyzed (Figure~\ref{fig:lang_main}). We restrict 
this analysis to programming languages with at least $20$ data points, as this 
is a requirement for our hypothesis tests. Thus, we compare the results for 
\texttt{C/C++}, \texttt{Ruby}, \texttt{Java}, \texttt{Objective C/C++}, 
\texttt{Python} and \texttt{PHP}. \texttt{Groovy} samples are not presented
because the number of cases did not fulfill the statistical test requirement. 
\texttt{C/C++} ($46.5\%$), \texttt{Ruby} ($46.5\%$) and 
\texttt{PHP} ($38.6\%$) are the programming languages with worse 
impact on software maintainability. \texttt{Java} ($50.9\%$) and \texttt{Python} 
($46.5\%$) seem to be the 
languages less affected by patching, i.e., more than $45\%$ of patches had a positive 
impact on maintainability. But overall patches in all languages have a 
considerable negative impact on maintainability---between $35\%$ to $50\%$---which 
confirms the need for better and more secure programming languages. 
Statistical significance was only retrieved for the \texttt{C/C++} 
(p-value = $2.24$x$10^{-05}$), \texttt{Ruby} (p-value = $0.041$) and \texttt{PHP} (p-value = $0.048$) languages.
The remaining programming languages did not yield statistical significance, 
thus results should be read with care.
Yet, data reports very interesting hints on the impact of programming languages 
on security patches.

We would expect that the negative impact of programming languages on
maintainability to be more severe, as arguably poor design of programming
languages and lack of using the best practices lead to more buggy and vulnerable
code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Ruby} have 
a significant negative impact approximate to $50\%$ on software 
maintainability. We suspect that these values are
the result of projects contributions policies (e.g., coding standards). In our dataset, 
$9$/$10$ projects with more contributors have strict contribution 
policies regarding code standards. For
instance, \texttt{laravel/framework} requires the \texttt{PSR-2
Standard}\footnote{PSR-2 Standard available at
\url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.

\textbf{Conclusion:} Results show that developers fail to follow the coding
practice guidelines and hinder software maintainability after patching; and, 
that different levels of attention should be paid to each guideline. For instance,
\emph{Write Simple Units of Code} and \emph{Write Short Units of Code}
are the most affected ones. No statistical significance was retrieved
for \emph{Separate Concerns in Modules}. 
The resulting p-value of the Paired Wilcoxon signed-rank test for $M(v)$ in Figure~\ref{fig:guidelines}
is $0.044$. Since the p-value is 
below the significance level of $0.05$, we argue that \emph{security patches 
may have a negative impact on the maintainability of open-source software}; and, 
that tools such as Better Code Hub should be integrated in the CI/CD pipeline
to help developers evaluate the risk of patches hindering software maintainability
alongside with Pull Requests/Code Reviews. We also conclude that different severity 
vulnerabilities may need different levels of attention and better and more secure 
programming languages are needed.
%
\\\textit{\textbf{RQ2: Which weaknesses are more likely to
affect open-source software maintainability?}}

@@

\section{Study Implications}\label{sec:implications}
%
Our results show evidence that developers may have to reduce maintainability for the 
sake of security. We argue that developers should be able to patch and produce secure 
code without hindering the maintainability of their projects. But there are still  
concerns that need to be addressed and that this study brings awareness for:
%
\\\textit{\textbf{Tools for Patch Risk Detection:}}
	Design debt of one guideline can lead
	to severe impacts on the software quality~\cite{10.1145/1985362.1985366}. 
	Thus, tools like Better Code Hub and static analysis (e.g., SonarQube, Codacy,
	ESLint, Infer, and more) should be leveraged to 
	assess the risk of patching vulnerabilities---alongside pull
  requests/code reviews. \Rui{what? (till the ned)-->} While static code analysis
	may be daunting due to the number of rules and different effects
	on maintainability, BCH uses the top-$10$ of the guidelines
	with the highest effect on maintainability. Yet, static analysis 
	can be used to complement the source code analysis, introduce
	the capability of vulnerability detection, and, predict and
  reduce the risk of hindering maintainability.
  \Rui{static analysis vs. source coode?!}
%
\\\textit{\textbf{Computer Curricula Needs to be Updated:}}
	We argue that software Developers lack the knowledge to create both secure and
	maintainable code. Mainly because academic curricula is not yet prepared
	to properly educate them. Universities should do their part on making
	this possible. \Rui{connection to curricula is not clear-->} Online services such as BCH have a great potential to help
	developers to be more aware of the maintainability issues introduced by their
	changes. Thus, developers can 
	put more work into improving their code maintainability and avoiding the risk
	of introducing new vulnerabilities while patching older ones.
%	
\\\textit{\textbf{Secure Programming Languages Wanted:}} Programming Languages 
	should provide new design patterns to easily implement security weaknesses 
	without endangering maintainability.
	Ultimately, new programming languages, both secure and maintainable by design,
	may be designed to help developers be highly productive when writing and
	maintaining secure applications. One example is the need for designing new
	authorization mechanisms since these are one of the most complex security
	features to implement.
%
\\\textit{\textbf{Poor Coding Practices Documentation:}}
  \Rui{too generic?}
	Software documentation must be provided to developers with the
	best practices to implement secure and maintainable software. Unmaintainable
	code is difficult to test, analyze and re-use. Software companies must
	do an effort to provide these artifacts to their developers, in order,
	to avoid the introduction of new vulnerabilities while patching older ones.

>>>>

\begin{abstract}
Security is a requirement of utmost importance to produce software with 
high quality. However, there is still a considerable amount of 
vulnerabilities being discovered and fixed, almost weekly. We suspect 
that when developers address these vulnerabilities on their codebases, 
they actually affect the maintainability of software, potentially 
introducing other vulnerabilities. This paper evaluates the impact of 
patches to improve security on the maintainability of open-source software. 
Maintainability is measured based on the Better Code Hub’s model of 10 
guidelines using a dataset including 1335 security commits.
Results show evidence of a new trade-off between security and maintainability, i.e., 
when developers may be hindering the software maintainability or even introducing 
new vulnerabilities while patching vulnerabiliteis.
We conclude that changes to codebases while patching 
vulnerabilities need to be performed with extra care (e.g., using tools
for patch risk assessment), computer science
curricula needs to be updated and more secure programming languages are necessary.
\end{abstract}
\begin{IEEEkeywords}
Security, Software Maintenance, Open-Source Software
\end{IEEEkeywords}

@@

 \begin{figure}[htp]
 	\vspace{-0.9cm}
     \begin{adjustwidth}{-1cm}{-1cm}  
  	\centering
  	\includegraphics[width=0.51\textwidth]{figures/main_per_guideline.pdf}
    \end{adjustwidth}
  	\caption{Impact of the security patches per guideline. For each guideline, $3$ types of impact 
	in the software maintainability are presented: negative, maintainability decreases (red); none, 
	maintainability remains the same (yellow); and, positive, maintainability increases (green).}
 	\label{fig:guidelines}	
	\vspace{-0.5cm}
 \end{figure}

Regarding the impact of security patches for each guideline, we observe
that overall patches have more cases where the impact is positive. However, 
we also see that patching vulnerabilities also has a very
significant number of negative cases for all guidelines---between $10\%$ and $40\%$. 
\emph{Write Short Units of Code} ($38.3\%$), \emph{Write Simple Units of Code} 
($37.9\%$), and \emph{Separate Concerns in Modules} ($33.8\%$) seem to be the 
most negativelly affected guidelines. This may imply that developers when patching 
vulnerabilities have a hard time to design/implement solutions that continue to 
respect the limit bounds of branch points and function/module sizes that are 
recommended by coding best practices. Still, on 
respecting bound limits, developers also seem to not consider the limit 
of $4$ parameters per function for the \emph{Keep Unit Interfaces Small} 
guideline required by BCH, in $24.8\%$ of the cases. This guideline is usually
violated when the patch requires to input new information to a function/class 
and developers struggle to use the \emph{Introduce Parameter Object} refactoring 
pattern. Results do not provide statistical significance to the \emph{Separate 
Concerns in Modules} guideline, i.e., results should be read carefully.

Software architecture is also affected while patching vulnerabilities.
Both \emph{Couple Architecture Component Loosely} and \emph{Keep
Architecture Components Balanced} guidelines suffer a negative impact of 
$26.9\%$ and $11.0\%$, respectively. Component independence and balance
are important to make easier to find the source code that developers
want to patch/improve and to understand how the high-level components
interact with others. However, results may imply that developers
forget to use techniques such as encapsulation to hide implementation
details and make the system more modular.

The \emph{Write Code Once} guideline results show that duplicated code 
increased in $171$ ($17.7\%$) of the patches. Software systems typically have 
$9\%$-$17\%$ of cloned code~\cite{5773403}. Previous work showed a 
correlation between code smells and code duplication~\cite{7476787} 
which may also be reflected in the \emph{Write Clean Code} guideline results. 
BCH reported new code smells for $145$ ($15.0\%$) patches which according 
to previous work may harm software security since some of these new code
smells may be new vulnerabilities~\cite{8819456}.
Thus, while developers 
tend to ignore refactoring techinques such as the \emph{Replace Method 
with Method Object} technique, they may also be introducing new software 
vulnerabilities and, consequently, harm the market value and economy of companies
~\cite{4267025}.

\textbf{Overall Score ($M(v)$):} 
Although overall patching vulnerabilities has a less negative impact
on software maintainability guidelines, this is not reflected in the average 
impact of all guidelines ($M(v)$) as we can see in Figure~\ref{fig:guidelines}.
The $M(v)$ plots show that $406$ ($41.9\%$) cases have negative impact
on software maintainability. While $188$ ($19.4\%$) cases have no impact, $375$ 
($38.7\%$) have a negative impact on software maintanibility. The larger number of 
negative cases may be explained by guidelines with higher concentration of negative 
cases with higher amplitudes, such as, \emph{Write Short Units of Code}, \emph{Write 
Simple Units of Code} and \emph{Separate Concerns in Modules}.

@@

\textbf{Severity:} Some of the vulnerabilities are identified with \emph{Common 
Vulnerabilities and Exposure} (CVE) entries. We leveraged the \emph{National 
Vulnerability Database} (NVD) website to collect their severity levels. In total, 
we retrieved severity scores for $536$ vulnerabilities: $112$ \emph{High}, 
$395$ \emph{Medium} and $29$ \emph{Low}. Figure~\ref{fig:severity} 
presents the impact of security patches per severity level on the 
maintainability of open-source software. We observe that patches for 
\emph{High} ($50.0\%$) and \emph{Medium} ($43.8\%$) severity vulnerabilities 
hinder more the maintainability of software than \emph{Low} ($27.6\%$) 
severity vulnerabilities. Again, patches have a considerable 
negative impact on software maintainability---between $20\%$ 
and $50\%$. Statistical significance was retrieved only for \emph{Low}
severity vulnerabilitie, i.e., \emph{Low} severity vulnerabilities
may have less negative impact on software maintainability thant the 
other severity levels. However, results should not be disregarded 
because they seem to confirm the 
assumption that higher severity vulnerabilities patches may have a 
more negative impact on maintainability and, consequently, that high/medium 
severity vulnerabilities may need more attention than low severity while 
patching vulnerabilities.

@@

\textbf{Programming Language:} The impact on software maintainability per programming 
language was also analyzed (Figure~\ref{fig:lang_main}). We restrict 
this analysis to programming languages with at least $20$ data points, as this 
is a requirement for our hypothesis tests. Thus, we compare the results for 
\texttt{C/C++}, \texttt{Ruby}, \texttt{Java}, \texttt{Objective C/C++}, 
\texttt{Python} and \texttt{PHP}, leaving \texttt{Groovy} out of the analysis.
\texttt{C/C++}, \texttt{Ruby} and 
\texttt{PHP} are the programming languages with worse 
impact on software maintainability, i.e., with the highest number of negative cases (respectively, $46.5\%$, $46.5\%$ and $38.6\%$). \texttt{Java} and \texttt{Python} 
seem to be the languages less affected by patching, i.e., integrating a larger amount of cases with positive impact on maintainability (respectively, $50.9\%$ and $46.5\%$). But overall 
languages have a considerable amount of cases that negatively impact software 
maintainability---between $35\%$ to $50\%$---which 
confirms the need for better and more secure programming languages. 
Statistical significance was only retrieved for the \texttt{C/C++} 
(p-value = $2.24$x$10^{-05}$), \texttt{Ruby} (p-value = $0.041$) and \texttt{PHP} (p-value = $0.048$) languages.
The remaining programming languages did not yield statistical significance, 
thus results should be read with care.
Yet, data reports very interesting hints on the impact of programming languages 
on security patches.

We expected the negative impact for programming languages on
maintainability to be more severe, as arguably, poor design of programming
languages for security and the lack of best practices application by developers   leads to more buggy and vulnerable
code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Ruby} have 
a significant negative impact approximate to $50\%$ on software 
maintainability. We suspect that these values are
the result of projects contributions policies (e.g., coding standards). In our dataset, 
$9$/$10$ projects with more contributors have strict contribution 
policies regarding code standards. For
instance, \texttt{laravel/framework} requires the \texttt{PSR-2
Standard}\footnote{PSR-2 Standard available at
\url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.

\textbf{Conclusion:} Results show that developers have a hard time followings the coding
practice guidelines and, consequently, hinder software maintainability while patching; and, 
that different levels of attention should be paid to each guideline. For instance,
\emph{Write Simple Units of Code} and \emph{Write Short Units of Code}
are the most affected ones. No statistical significance was retrieved
for \emph{Separate Concerns in Modules}. 
The resulting p-value of the Paired Wilcoxon signed-rank test for $M(v)$ in Figure~\ref{fig:guidelines}
is $0.044$. Since the p-value is 
below the significance level of $0.05$, we argue that \emph{security patches 
may have a negative impact on the maintainability of open-source software}; and, 
that tools such as Better Code Hub should be integrated in the CI/CD pipeline
to help developers evaluate the risk of patches hindering software maintainability
alongside with Pull Requests/Code Reviews. We also conclude that different severity 
vulnerabilities may need different levels of attention and better and more secure 
programming languages are needed.
%
\\\textit{\textbf{RQ2: Which weaknesses are more likely to
affect open-source software maintainability?}}

@@

\section{Study Implications}\label{sec:implications}
%
Our results show evidence that developers may have to reduce maintainability for the 
sake of security. We argue that developers should be able to patch and produce secure 
code without hindering the maintainability of their projects. But there are still  
concerns that need to be addressed and that this study brings awareness for:
%
\\\textit{\textbf{Tools for Patch Risk Assessment:}}
	Design debt of one guideline can lead
	to severe impacts on the software quality~\cite{10.1145/1985362.1985366}. 
	Thus, tools like Better Code Hub and static analysis (e.g., SonarQube, Codacy,
	ESLint, Infer, and more) should be leveraged to 
	assess the risk of patching vulnerabilities---alongside pull
  requests/code reviews. Static code analysis
	may be daunting due to the number of rules and different effects
	on maintainability. BCH claims to use the top-$10$ of the guidelines
	with the highest effect on maintainability. Yet, static analysis 
	can be used to complement the BCH analysis by introducing
	the capability of vulnerability detection and support the prediction/prevention of the risk of hindering maintainability while patching.
%
\\\textit{\textbf{Computer Science Curricula Needs to be Updated:}}
	We argue that developers lack the knowledge to create both secure and
	maintainable code. Mainly because academic curricula is not yet prepared
	to properly educate them. Universities should encourage students to use online 	services such as BCH or similar ones. These tools have a great potential to make
	students aware of maintainability issues and beginner mistakes (e.g.,
	coding practices violation). Thus, students and future developers can 
	put more work into improving their code maintainability and avoiding the risk
	of introducing new vulnerabilities while patching older ones.
%	
\\\textit{\textbf{Secure Programming Languages Wanted:}} Programming Languages 
	should provide new design patterns to easily implement security weaknesses 
	without endangering maintainability.
	Ultimately, new programming languages, both secure and maintainable by design,
	may be designed to help developers be highly productive when writing and
	maintaining secure applications. One example is the need for designing new
	authorization mechanisms since these are one of the most complex security
	features to implement.
% %
% \\\textit{\textbf{Poor Coding Practices Documentation:}}
%   \Rui{too generic?}
% 	Software documentation must be provided to developers with the
% 	best practices to implement secure and maintainable software. Unmaintainable
% 	code is difficult to test, analyze and re-use. Software companies must
% 	do an effort to provide these artifacts to their developers, in order,
% 	to avoid the introduction of new vulnerabilities while patching older ones.
